---
title: "Introduction to R"
subtitle: "R for Public Health"
author: "Ashwini Kalantri"
date: "2023-12-18"
date-format: DD MMM YYYY
logo: r23_head.png
institute: Department of Community Medicine, MGIMS
format:
  revealjs:
    theme: night
    slide-number: true
    css: styles.css
---

# Get started
```{r}
#| echo: true
#| output-location: column-fragment
"Hello World!"
```

## Assignment {auto-animate="true"}
`<-`
```{r}
#| echo: true
#| output-location: column-fragment
text1 <- "Hello World!"
print(text1)
```
`->`
```{r}
#| echo: true
#| output-location: column-fragment
"Hello World!" -> text2
print(text2)
```
`=`
```{r}
#| echo: true
#| output-location: column-fragment
text3 = "Hello World!"
print(text3)
```


## Reserved Words {auto-animate="true"}
:::: {.columns}

::: {.column width="30%"}
- if
- else
- while
- repeat
- for
- function
- in
:::

::: {.column width="30%"}
- next
- break
- TRUE
- FALSE
- NULL
- Inf
- NaN
:::

::: {.column width="30%"}
- NA
- NA_integer
- NA_real
- NA_complex_
- NA_character_
- …
:::

::::


## Commenting {auto-animate="true"}

```{r}
#| echo: true
#| output-location: fragment

# Print the text "Hello World"
print("Hello World!")
```

```{r}
#| echo: true
#| output-location: fragment

print("Hello World!") # Print the text "Hello World"
```

```{r}
#| echo: true
#| output-location: fragment

# Multi-line comment
# about printing the text "Hello World"
print("Hello World!")
```

# Operators

## Arithmetic {auto-animate="true"}
Addition
```{r}
#| echo: true
#| output-location: column-fragment
2 + 5
```

Subtraction
```{r}
#| echo: true
#| output-location: column-fragment
73 - 32
```

Multiplication
```{r}
#| echo: true
#| output-location: column-fragment
47 * 7
```

Division
```{r}
#| echo: true
#| output-location: column-fragment
86 / 3
```

## Arithmetic {auto-animate="true" visibility="uncounted"}
Exponentiation
```{r}
#| echo: true
#| output-location: column-fragment
8^2
```

Modulus
```{r}
#| echo: true
#| output-location: column-fragment
77%%3

```

## Relational  {auto-animate="true"}
Greater
```{r}
#| echo: true
#| output-location: column-fragment
5 > 6
```
Lesser
```{r}
#| echo: true
#| output-location: column-fragment
5 < 6
```
Equal
```{r}
#| echo: true
#| output-location: column-fragment
6 == 6
```
## Relational  {auto-animate="true" visibility="uncounted"}
Greater or equal
```{r}
#| echo: true
#| output-location: column-fragment
8 >= 5
```
Lesser or equal
```{r}
#| echo: true
#| output-location: column-fragment
7 <= 10
```
Not Equal
```{r}
#| echo: true
#| output-location: column-fragment
9 != 10
```

## Joining Logical 
`AND`
```{r}
#| echo: true
#| output-location: column-fragment
TRUE & TRUE
```

```{r}
#| echo: true
#| output-location: column-fragment
TRUE & FALSE
```

```{r}
#| echo: true
#| output-location: column-fragment
FALSE & FALSE
```

`OR`
```{r}
#| echo: true
#| output-location: column-fragment
TRUE | TRUE
```

```{r}
#| echo: true
#| output-location: column-fragment
TRUE | FALSE
```

```{r}
#| echo: true
#| output-location: column-fragment
FALSE | FALSE
```

## Classes {auto-animate="true"}

Integer
```{r}
#| echo: true
#| output-location: column-fragment
int <- 3L
print(int)
```

```{r}
#| echo: true
#| output-location: column-fragment
class(int)
```
Numeric
```{r}
#| echo: true
#| output-location: column-fragment
num <- 4.3
print(num)
```

```{r}
#| echo: true
#| output-location: column-fragment
class(num)
```
Character
```{r}
#| echo: true
#| output-location: column-fragment
name <- "Your Name"
print(name)
```

```{r}
#| echo: true
#| output-location: column-fragment
class(name)
```

## Classes {auto-animate="true" visibility="uncounted"}
Logical
```{r}
#| echo: true
#| output-location: column-fragment
logT <- TRUE
logF <- F
print(logF)
```

```{r}
#| echo: true
#| output-location: column-fragment
class(logF)
```
Date
```{r}
#| echo: true
#| output-location: fragment
date1 <- "2023-12-18"
date2 <- 2023-12-18
date3 <- as.Date("2023-12-18")
date4 <- as.Date("18 Dec 2023","%d %b %Y")
date5 <- as.Date(45076, origin = "1900-01-01")
print(date1)
```

```{r}
#| echo: true
#| output-location: column-fragment
class(date1)
```

## Class Conversion {auto-animate="true}
```{r}
#| echo: true
num <- "1"
num <- as.numeric(num)

numLet <- as.numeric(LETTERS)
charNum <- as.character(1:100)

tf <- c("TRUE","FALSE","FALSE")
tf <- as.logical(tf)

num <- as.character(num)
```

# Objects {auto-animate="true" visibility="uncounted"}

## Vectors {auto-animate="true"}
```{r}
#| echo: true
#| output-location: fragment

vec1 <- c(2,4,6,8,3,5.5)
vec2 <- 4

#combining vectors
newVec <- c(vec1,vec2)

newVec
```

```{r}
#| echo: true
#| output-location: fragment
dateVec <- c(as.Date("2023-11-28"),
             as.Date("2023-12-22"),
             Sys.Date())
dateVec
```

```{r}
#| echo: true
#| output-location: fragment
newVec[5]
```

## Matrix {auto-animate="true"}
```{r}
#| echo: true
#| output-location: column-fragment
let <- matrix(LETTERS,
              nrow = 6,
              ncol = 6,
              byrow = F)

let
```

```{r}
#| echo: true
#| output-location: column-fragment
let[3,5]
```

```{r}
#| echo: true
#| output-location: column-fragment
let[,5]
```

```{r}
#| echo: true
#| output-location: column-fragment
let[5,]
```

## Factor {auto-animate="true"}
```{r}
#| echo: true
#| output-location: fragment
gender <- c(1,2,2,1,1,1,2,2,1,2,1)

genFac <- factor(gender,
                 levels = c(1,2),
                 labels = c("Male","Female"))

genFac
```

## Data Frames {auto-animate="true"}

```{r}
#| echo: true
age <- c(12,24,NA,23,65,33) # create age vector

gender <- c("M","F","F","M","M","F") #create gender vector

occu <- factor(c(1,4,3,2,4,5), #occupation 
               levels = c(1:5),
               labels = c("Unemp","Service","Student","Business","Prof"))

#date of birth
dob <- c(as.Date("1992-01-11"),as.Date("1992-12-11"),as.Date("1992-11-15"),
         as.Date("1982-11-11"),as.Date("1992-05-11"),as.Date("1992-11-21"))

#create data frame
df <- data.frame(age,gender,occu,dob)
```

## Data Frames {auto-animate="true" visibility="uncounted"}
```{r}
#| echo: true
#| output-location: column-fragment
df
```

```{r}
#| echo: true
#| output-location: column-fragment
df[2,]
```

```{r}
#| echo: true
#| output-location: column-fragment
df[,2]
```

```{r}
#| echo: true
#| output-location: column-fragment
df[2]
```

```{r}
#| echo: true
#| output-location: column-fragment
df[2,4]
```

## List {auto-animate="true"}
```{r}
#| echo: true
#| output-location: column-fragment
list <- list(df,dob,let,newVec)

list
```

## List {auto-animate="true" visibility="uncounted"}
List with nth object(s)
```{r}
#| echo: true
#| output-location: column-fragment
list[2]
```

nth object
```{r}
#| echo: true
#| output-location: column-fragment
list[[2]]
```
selecting withing object
```{r}
#| echo: true
#| output-location: column-fragment
list[[2]][4]
```

```{r}
#| echo: true
#| output-location: column-fragment
list[[1]][2,3]
```


# Functions {auto-animate="true"}
```{r}
#| echo: true
#| output-location: column-fragment
addition <- function(n1,n2){
  n1 + n2
}

div <- function(n1,n2){
  n1 / n2
}
```

## Functions {auto-animate="true"}
```{r}
#| echo: true
#| output-location: column-fragment

div(n1 = 55,n2 = 3)
```

# Packages {auto-animate="true"}
```{r}
#| echo: true
#| output-location: column-fragment
library(dplyr)

dplyr::glimpse(df)
```
# Working Directory {auto-animate="true"}
```{.r}
setwd("~/r4ph23")

setwd("C:/user/ashwini/documents/r4ph23")

getwd()
```

# Importing Data {auto-animate="true"}

## CSV {auto-animate="true"  visibility="uncounted"}
```{r}
#| echo: true
data <- read.csv("data.csv")
```

## Excel {auto-animate="true"  visibility="uncounted"}
```{r}
#| echo: true
library(readxl)
data <- read_excel("data.xlsx")
```

## Stata, SPSS {auto-animate="true"  visibility="uncounted"}
```{r}
#| echo: true
library(haven)
data <- read_sav("data.sav")
data <- read_dta("data.dta")
```

# Getting Help {auto-animate="true"}

## Help  yourself {auto-animate="true"  visibility="uncounted"}

- Read the manual
- Check your code
- Read the error message
- Web search
- Read the forums
- Rubber duck debugging

## Get Help {auto-animate="true"  visibility="uncounted"}

- Ask a friend, colleague 
- Post of the forums
  - Describe your goal
  - Be explicit about your question
  - Provide specific information
  - Be courteous
  - Provide the solution if you found it elsewhere.